Research Notes

### 2018-10-30 writeups/notes.tex
huh

\end{document}

>>>>

\section{Hydraulic Jumps in Surface Waves}

We originally got started on this since it seems like a dimensionally-restricted
version of what we're seeing in simulations, but presently unclear whether
really the case. Also, Faber's book for fluids \emph{Fluid Dynamics for
Physicists} cites a critical breaking jump height of $1.3$ which is markedly
false; papers such as Chanson's 2009 hydraulic jumps review in Euro.\ J.\ Mech\
B/Fluids shows that undular jumps (instead of turbulent jumps) can be found
easily up to $3.5/4$.

\subsection{1D Hydraulic Jump}

This is one of the OG papers; B\'elanger solved this as early as in 1838 but
Rayleigh's work in 1916 is also pretty famous. Consider the 1D nonlinear Euler
equations for a body of water whose surface can move $h(x)$, then give it
velocity $u(x, z)$. The problem is really 2D, but we can integrate along the $z$
direction, in which case $\pd{P(x, z)}{x} = \frac{\rho g h(x, z)}{x}$ simply
integrates to $\frac{\rho g h(x)^2}{2} = \frac{g\Sigma(x)^2}{2\rho}$. Assuming
also the velocity is uniform $u(x, z) = u(x)$ gives equations of motion
\begin{align}
    \pd{\Sigma}{t} + \pd{(\Sigma u)}{x} &= 0,\\
    \pd{u}{t} + u\pd{u}{x} &=
        -\frac{1}{\Sigma}\pd{}{x}\p*{\frac{g\Sigma^2}{2\rho}},\nonumber\\
        &= -\frac{g}{\rho}\pd{\Sigma}{x}.
\end{align}
It proves easier to divide the density equation by $\rho$ so that we can replace
$\frac{\Sigma}{\rho} = h(x)$ in the above. A specific momentum conservation
equation can be obtained
\begin{equation}
    \pd{(hu)}{t} + \pd{}{x}\p*{hu^2 + \frac{gh^2}{2}} = 0.
\end{equation}

Now, consider a hydraulic jump $u_1, h_1 \to u_2, h_2$ such that the
discontinuity is fixed (we've shifted to the jump front frame), so that
$\s*{uh}, \s*{hu^2 + gh^2/2}$ bracketed quantities are continuous across the jump.
This results in
\begin{align}
    h_1u_1^2 - h_2u_2^2 = h_1u_1^2\p*{1 - \frac{u_2}{u_1}}
        &= h_1u_1^2\p*{1 - \frac{h_1}{h_2}},\nonumber\\
        &= \frac{g\p*{h_2^2 - h_1^2}}{2},\nonumber\\
    u_1^2 &= \frac{h_2}{h_1}\frac{g\p*{h_1 + h_2}}{2}.
\end{align}
Customarily one defines $a_1^2 = gh_1, \mathrm{Fr} =
\frac{u_1}{a_1}$ and $\gamma = \frac{h_2}{h_1}$ which gives simplification
\begin{align}
    2\mathrm{Fr}^2 &= \gamma(1 + \gamma)\nonumber\\
    \gamma = \frac{-1 + \sqrt{1 + 8\mathrm{Fr}^2}}{2}.
\end{align}
Note now that $\mathrm{Fr}^2 > 1$ produces a $\gamma > 1$, so $h_2 > h_1$.
Noting moreover that $\frac{u_1}{u_2} = \gamma$ and so $u_1 > u_2$. If we have a
picture of a jet incident on a medium at rest, $u_2$ is just the velocity of the
jump.

\subsection{1D Viscous Theory}

We follow Bohr's paper from J.\ Fluid Mech.\ in 1993 (also done by Tani in 1949
to some extent) but in 1D first so that the equations are a bit easier to
intuit. It's completely unnecessary but illustrates some features of the
approach.

Consider the same 2D equations as before (before integrating along $z$), but now
permit a viscosity term. In the boundary layer approximation, the usual
assumptions are that flow largely varies horizontally but variations are
stronger vertically (i.e.\ we only consider the $x$ velocity equation but only
consider the $z$ viscosity). We make one small modification, it seems that we
permit an incompressibility assumption since $\rd{h(x, z, t)}{t} = 0$; this is
just $\Delta P = \rho g h = $ the Lagrangian pressure boundary condition. Thus,
the equations of motion are
\begin{align}
    \pd{u_x}{x} + \pd{u_z}{z} &= 0,\\
    u_x \pd{u_x}{x} + u_z \pd{u_x}{z} &= -g\pd{h}{x} + \nu \ptd{u_x}{z}.
\end{align}
Integration along $z$ of the momentum equation can be done with appropriate BCs:
no-slip/penetration at the bottom $u_x(z = 0) = u_z(z = 0) = 0$, while no shear
at the surface $\pd{u_x}{z}(z = h) = 0$. We handle the integral of $u_z u_{x,
z}$ by integrating by parts, dropping the boundary term and continuity equation,
giving
\begin{equation}
    2u_x \int\limits_0^h u_x\rd{u_x}{x}\;\mathrm{d}z
        = -g\pd{h}{x}h - \nu\at{\pd{u_x}{z}}_{z = 0}.
\end{equation}
We make a few handwavy substitutions for representative values $\partial_z u_x(z
= 0) \sim u_x / h$ and $\partial_x u_x^2 \sim \partial_x \bar{u}_x^2$ average
value. This averaging procedure and dropping the factor of $2$ (the two papers
seem to disagree on this coefficient, I think Bohr's paper has an error) lets
us examine
\begin{equation}
    \bar{u}_x \pd{\bar{u}_x}{x} + g\pd{h}{x} \sim -\frac{\bar{u}_x}{h^2}.
\end{equation}
Next, let's identify $\bar{u}_xh \equiv q$ volume flux constant in $x$, which
leaves us with
\begin{equation}
    \pd{\bar{u}_x}{x}\p*{\bar{u}_x - \frac{gq}{\bar{u}_x^2}}
        \sim -\nu\frac{\bar{u}_x^3}{q^2}.
\end{equation}
Let's study in terms of $\mathrm{Fr}^2 = \frac{\bar{u}_x^2}{gh} =
\frac{\bar{u}_x^3}{gq}$ again, albeit a bit messier, which takes a bit of work
but eventually lets us rearrange
\begin{equation}
    \pd{\mathrm{Fr}}{x} = \frac{\p*{gq}^{1/3}\nu}{q^2}\frac{
        \mathrm{Fr}^{5/3}}{1 - \mathrm{Fr}^2}.
\end{equation}
Note that the units work here, as $\frac{\p*{gq}^{1/3}\nu}{q^2} \equiv L_\nu$
has units of length.

Let's interpret the equation we have now: we have exactly how the inbound
$\mathrm{Fr}\p*{x = -\infty}$ evolves over $x$, which should describe (extremely
coarsely, owing to our approximations) the boundary layer of the jump.

Qualitatively, we can use a parameterization $\mathrm{Fr}(s), x(s)$ such that
$\dot{\mathrm{Fr}}(s) = -\mathrm{Fr}^{5/3}, \dot{x}(s) = L_\nu \p*{\mathrm{Fr}^2
- 1}$ where we define $L_\nu \equiv a_0h^2/\nu$ and we've absorbed the negative
sign into $\dot{\mathrm{Fr}}(s)$. I'm too lazy to plot this, but if we start at
a point $\mathrm{Fr} > 1$, it's clear that $\mathrm{Fr}$ increases and $x$
decreases until $\mathrm{Fr} = 1$. Since the sign convention on the parametric
is arbitrary, we can flip the sign and see that the parameterized curve formally
follows decreasing $\dot{\mathrm{Fr}}(s) \propto -1/s$ while $\dot{x}(s) \propto
-L$. This is double-valued with the solution $\mathrm{Fr} > 1$ though, so with
some physical intuition we discard this and consider what values of
$\mathrm{Fr}$ can take on beyond its intersection with $\mathrm{Fr} = 1$.

It must discontinuously jump to another family of curves. But any nonzero value
of $\mathrm{Fr}$ will live on another one of these such trajectories and
increase without bound; indeed the only reasonable asymptotic solution is
$\mathrm{Fr} = 0$. This completes the picture: for an inbound $\mathrm{Fr} \gg
1$, it follows roughly $\pd{\mathrm{Fr}}{x} \gtrsim -\mathrm{Fr}^{-1/3}$ until
it reaches $\mathrm{Fr} = 1$, then it discontinously disappears. This is the
upstream flow, and so by mass conservation the location of the discontinuity
must travel.

This is exactly the picture we derived above: for an incoming $\mathrm{Fr}
\gtrsim 1$, the upstream flow has zero velocity and the jump travels.

\subsection{2D Hydraulic Jump}

This is the case worked out in detail in Bohr's paper. Following much the same
approach but in cylindrical coordinates we may write down equations of motion
\begin{align}
    u_r \pd{u_r}{r} + u_z \pd{u_r}{z} &= -g\pd{h}{r} + \nu
        \ptd{u_r}{z},\nonumber\\
    \frac{1}{r}\pd{(ru_r)}{r} + \pd{u_z}{z} &= 0.
\end{align}
Note the slight difference here due to the $\vec{\nabla} \cdot$ in cylindrical
coordinates. Interestingly, as a consequence of this $\pd{P_r}{t} +
\frac{1}{r}\partial_r \p*{hu_r^2 + gh^2/2} \neq 0$, since one term is a
$\vec{\nabla} \cdot (hu_r)$ and the other is a $\vec{\nabla}gh^2/2$! Radial
momentum is not conserved as the flow spreads out. Averaging using the same
techniques as above gives
\begin{align}
    \bar{u}_r\pd{\bar{u}_r}{r} + g\pd{h}{r} &= -\nu \frac{\bar{u}_r}{h^2},\nonumber\\
    \bar{u}_rhr &= q.
\end{align}
Combining the two gives
\begin{align}
    \pd{\bar{u}_r}{r}\p*{\bar{u}_r - \frac{gq}{\bar{u}_r^2r}}
        &= \frac{gq}{\bar{u}_rr^2} - \nu\frac{\bar{u}_r^3r^2}{q^2}.
        \label{eq:2d_hjump}
\end{align}
It's not very worth converting into dimensionless $\mathrm{Fr}$ number since it
would carry an $r$ dependence (if we are to hold $q$ constant) and we need to
study the $r$ dependence.

It may be observed that \autoref{eq:2d_hjump} has only a single fixed point,
$u_r^3r = gq, u_r^4r^4 = gq^3/\nu$. Call this point $r_c$ the critical radius,
then it turns out that the two parametric solutions (as we analyzed in the
previous section) both spiral around this point. This has a few important
properties:
\begin{itemize}
    \item No asymptotic solutions: the solution for $r > r_c$ always diverges
        at some finite $r_s < \infty$. We can try an approximate solution in the
        form $u_r = y / r$; this is done in Bohr's paper.

    \item For a fixed $r_s \gg r_c, < \infty$, we can require the usual jump
        condition which takes the slightly modified form
\end{itemize}

Edit: Looks like Bhagat 2018 J.\ Fluid.\ Mech.\ does this more correctly, they
find that surface tension is the proper term to invoke after performing this
boundary layer analysis. I'm not going to finish reworking the rest of Bohr's
paper here, since it's up through the boundary layer equations that is reusable.
\end{document}
